Predicting the 2024 US Presidential Election with a Model-Based Forecast

### 2024-11-03 paper/paper.qmd
update model, paper, scripts

### Percent Support for Donald Trump
The primary outcome variable is the percentage (`pct`) of respondents who indicated support for Donald Trump in the 2024 general election. This percentage is calculated based on the total number of respondents in each poll divide by `sample_size`. The data includes observations from various states, providing both national and state-specific measures of Trump’s support.

@@

### Polling Methodology
The `methodology` variable describes the method each pollster used to collect data. It includes approaches such as live phone interviews, online surveys, and mixed methods. For polls that used multiple collection methods (e.g., phone and online), we labeled them as "Mixed Voting" for simplicity. The methodology is important because different approaches can introduce varying degrees of bias, influencing the final reported percentages.

@@

# Define the number of electoral votes for each state
electoral_seats <- data.frame(
  state = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida",
            "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", 
            "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", 
            "Missouri", "Montana", "Nebraska", "Nevada", 
            "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
            "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", 
            "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming", 
            "District of Columbia"),
  electoral_votes = c(9, 3, 11, 6, 55, 9, 7, 3, 29, 16, 4, 4, 20, 11, 6, 6, 8, 8, 4, 10, 11, 15, 10, 6, 10, 
                      3, 10, 6, 4, 14, 5, 29, 16, 3, 18, 7, 6, 20, 4, 9, 3, 11, 38, 6, 3, 13, 12, 5, 10, 3, 3)
)

@@

### Polling Score and Transparency
The `numeric_grade` variable is a rating assigned to each polling organization, indicating the reliability of the poll. Pollsters with higher numeric grades are considered more reliable and consistent in their methodology (e.g. higher than 2.5). Additionally, the `transparency_score` measures how openly a pollster shares their methodology and data, which further affects the trustworthiness of their results.

@@

- **Polling Scores**: Most organizations have numeric grades between 2.8 and 3.0, indicating that the majority of pollsters are considered highly reliable. Very few have grades below 2.7, which would suggest lower reliability.

- **Transparency Scores**: Transparency scores show more variability, with many pollsters achieving scores above 9, reflecting a high level of openness in their methodology. Fewer organizations have transparency scores below 6, indicating that while some pollsters are less open, the majority strive for transparency in their practices.

### Sample Size
The `sample_size` variable represents the total number of respondents surveyed in each poll. A larger sample size generally leads to more reliable and precise estimates of voter sentiment, as it reduces the margin of error. In contrast, smaller sample sizes can result in greater variability and less confidence in the results.

### Poll Duration
The `duration` variable is a constructed variable that represents how long Donald Trump has been in the 2024 presidential campaign, measured as the number of days from his official campaign announcement on July 15, 2024 until the `end_date` of each poll. This variable helps quantify how much time has passed since Trump officially entered the race and allows us to examine how his support has evolved over the course of his campaign.

# Model {#sec-model}
Our modeling approach aims to quantify the relationship between various polling metrics and the percentage support for a candidate. For this analysis, we use a linear model (LM) to examine how factors such as poll score, methodology, poll duration, sample size, transparency score, and state influence support percentages. The model is implemented using the lm function, with a Gaussian distribution to capture the variability in support rates.

@@

The model assumes that the distribution of support percentage, given these polling characteristics, follows a normal distribution. This Gaussian assumption facilitates parameter estimation, which is a standard approach in linear regression. To prevent overfitting and ensure interpretability, we assume moderate priors, maintaining balanced uncertainty across our predictors. This approach enables us to assess the impact of polling characteristics on candidate support while ensuring stability in our findings.

## Model set-up
The model predicts Trump's support percentage using the following predictor variables:

- Poll Score (`numeric_grade`): Represents the quality rating of the poll.

@@

- Poll Duration (`duration`): The the length of the election period was conducted.

- Sample Size (`sample_size`): The number of respondents in the poll.

- Transparency Score (`transparency_score`): A measure of how transparent the polling data and methodology are.

@@

\begin{align*}
    y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
    \mu_i &= \beta_0 + \beta_1 \cdot \text{Poll Score}_i + \beta_2 \cdot \text{Methodology}_i \\
          &\quad + \beta_3 \cdot \text{Poll Duration}_i + \beta_4 \cdot \text{Sample Size}_i \\
          &\quad + \beta_5 \cdot \text{Rransparency Score}_i + \beta_6 \cdot \text{State}_i + \epsilon_i \\
    \epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align*}

@@

### Model justification
Existing research and political science theories suggest that factors such as poll quality, sample size, transparency, methodology, poll duration, and the state where the poll takes place can influence Trump's support percentage. Higher quality polls and larger sample sizes are believed to yield more reliable estimates, while transparency scores can affect levels of public trust, potentially altering support. Moreover, different polling methodologies (e.g., online surveys, telephone interviews) may affect how representative the sample is, while longer polling durations can reflect shifts in public sentiment over time. State-specific factors, such as local political events or population characteristics, can also contribute to the variation in support. The timing of the poll, particularly in relation to key political events, might further influence public opinion.

@@

We predicted the outcome based on our test dataset. However, due to missing data for certain states in our original dataset, we were unable to make predictions for all states. This limitation arises because the states with NA values in the dataset are not represented in the model, preventing accurate forecasts for those regions.

## Model Results and Interpretation
The linear regression model built by our training dataset, using 384 data points, estimated the factors influencing Trump's support levels. As @tbl-summary-model shows, the intercept, estimated at 43.39, represents the level of support when all other predictors are at baseline. The model achieved an R-squared value of 0.85, indicating that 85% of the variance in Trump's support percentage can be explained by the predictors included in the model. The adjusted R-squared value of 0.82 also shows that the model effectively captures the relationships between variables without overfitting. The Root Mean Square Error (RMSE) was 2.03, indicating the average difference between predicted and actual values. The relatively low RMSE value suggests that the model provides reasonably accurate predictions.

The results highlight that `numeric_grade` and `sample_size` are particularly important in predicting support levels, with larger and higher-quality polls offering more reliable estimates. The transparency score and state-level coefficients further provide understanding into regional differences and the role of trust in polling, helping us better understand the dynamics behind Trump's public support.The coefficients for `numeric_grade` and `sample_size` underline the importance of poll quality and sample size in predicting support levels. Larger, higher-quality polls offer more estimates. Additionally, transparency and state effects provide insight into regional differences and trust factors affecting polling results, giving a clearer understanding of the dynamics influencing Trump's public support.

```{r}
#| label: tbl-summary-model
#| tbl-cap: "Summary of the Linear Regression Model for Predicting Polling Outcomes"
#| echo: false
#| warning: false

model2 <- readRDS(file = here::here("models/first_model.rds"))

modelsummary(
  list(
    "Linear Model" = model2
  ),
  statistic = NULL 
)
```

## Predicted Electoral Outcomes
We employed a regression model to estimate the percentage of votes Trump is likely to receive in each state. Using these predicted results and the electoral vote allocations, we identified the winner for each state and calculated the total electoral votes for both Trump and Harris.

@@

- In Arizona, Trump is expected to receive 48.94% of the vote, resulting in a win for Harris with 11 electoral votes.
- In Missouri, Trump is projected to receive 51.64%, thereby winning all 10 electoral votes for that state.
- In Ohio, with an estimated 51.98% of the vote, Trump secures Ohio’s 18 electoral votes.

Electoral Vote Count: Trump Electoral Votes: 98 - Harris Electoral Votes: 239. This indicates that Harris is projected to win the 2024 U.S. Presidential Election with 239 electoral votes, while Trump is projected to gather 98. These results hinge on the outcomes in various battleground states, with several being extremely close.

@@

# Define the number of electoral votes for each state
electoral_votes <- data.frame(
  state = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida",
            "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", 
            "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", 
            "Missouri", "Montana", "Nebraska", "Nevada", 
            "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
            "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", 
            "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming", 
            "District of Columbia"),
  electoral_votes = c(9, 3, 11, 6, 55, 9, 7, 3, 29, 16, 4, 4, 20, 11, 6, 6, 8, 8, 4, 10, 11, 15, 10, 6, 10, 
                      3, 10, 6, 4, 14, 5, 29, 16, 3, 18, 7, 6, 20, 4, 9, 3, 11, 38, 6, 3, 13, 12, 5, 10, 3, 3))

# Convert methodology column to character type for recoding
analysis_data_test$methodology <- as.character(analysis_data_test$methodology)

# Recode the 'Email' level in the methodology column to 'Online Panel'
analysis_data_test$methodology[analysis_data_test$methodology == "Email"] <- "Online Panel"

# Use the trained regression model to predict Trump's support percentage in each state
analysis_data_test$predicted_percent_trump <- predict(model2, newdata = analysis_data_test)
total_trump <- aggregate(predicted_percent_trump ~ state, data = analysis_data_test, FUN = mean)

@@

# Create and display the results table
kable(state_predictions[, c("state", "predicted_percent_trump", "electoral_votes", "winner")], 
      col.names = c("State", "Trump Predicted %", "Electoral Votes", "Winner"), 
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  scroll_box(width = "100%", height = "400px")
```

## Prediction map

```{r}
#| label: fig-statemap
#| fig-cap: Predicted Winner of 2024 US Presidential Election by State
#| echo: false
#| warning: false
#| message: false

# "red" for Trump and "blue" for Harris
state_predictions$winner_color <- ifelse(state_predictions$winner == "Trump", "red", "blue")

# Create a map visualization showing the predicted winner by state
plot_usmap(data = state_predictions, values = "winner_color", regions = "states") +
  scale_fill_manual(values = c("red" = "red", "blue" = "blue"), name = "Predicted Winner", labels = c("Harris", "Trump")) +
  labs(title = "Predicted Winner of the 2024 U.S. Presidential Election by State") +
  theme(legend.position = "right")

```
@fig-statemap displays the predicted outcome for each state in the 2024 U.S. Presidential Election, with green representing states where Trump is expected to win and blue for those where Harris is predicted to prevail.

The map illustrates Trump gaining traction across central and southern states, including Texas and Missouri, which have traditionally leaned Republican. Meanwhile, Harris holds a solid lead in states along the West Coast and in the Northeast, such as California and New York, regions that have a history of Democratic support.

Several battleground states, including Florida and Ohio, are projected to favor Trump, with a slight advantage that marks a contrast to Harris's stronger leads in places like California. The map also shows a number of central states in gray, reflecting either a lack of definitive prediction or insufficient data. Overall, the map reveals distinct regional patterns, with Trump expected to perform well in the heartland and Harris finding support along the coasts. The outcome may hinge on the final results in the competitive states like Pennsylvania, Arizona, and North Carolina.

@@

## Interpretation of Polling Data and Electoral Implications

This paper examines the factors affecting Donald Trump’s polling performance in the 2024 U.S. Presidential Election. Our linear regression model shows that poll duration and state-specific factors are the strongest predictors of Trump’s support. Longer poll durations tend to produce more stable data by capturing changes in public sentiment over time rather than short-term reactions. The state variable underscores the importance of local dynamics, as individual states bring unique policy priorities and demographic trends to their voter bases. These regional distinctions are essential in the Electoral College system, where state-level support directly impacts the final electoral count.

Transparency and sample size play secondary roles in data reliability. Polls with high transparency scores provide clear information about methodology, making them more reliable for analysis. Larger sample sizes reduce margins of error, adding stability to projections, especially in closely contested states. While both transparency and sample size support reliable predictions, they have less influence on voter preferences than duration and state-specific factors.

In comparison, polling methodology (e.g., live phone, online) and numeric grade (a general reliability rating) have minimal impact on Trump’s support in our model. Although these factors contribute to overall data quality, they do not significantly shape voter sentiment. This finding suggests that regional context and polling consistency over time are more important in understanding Trump’s polling performance in this election than the specific techniques or reputations of polling organizations.

## Influence of State Policies on Voter Preferences

Beyond the mechanics of polling, our analysis suggests that state-specific policies and regional cultural factors could influence voter preferences and ultimately shape electoral outcomes. For instance, certain policies, such as the legalization of recreational marijuana, reflect broader ideological values that may correlate with support for particular candidates. Although our model does not directly analyze the impact of marijuana policies, including these cultural aspects could offer a more comprehensive view of voter alignment.

States with legal recreational marijuana, for example, might lean more progressive or libertarian, potentially aligning with candidates like Harris who advocate for policies seen as liberal or socially progressive. This policy environment might signal broader state values that align with a particular party's stance, beyond just the specific candidate. Conversely, states that prohibit recreational marijuana use may lean more conservative, suggesting they may have stronger support for Trump. Although this relationship is not absolute, understanding such cultural factors helps contextualize the state-level support each candidate may receive and could be valuable in identifying battleground states. Additionally, factors like local economic policies, educational investment, or healthcare reform could influence voter preferences, making these areas ripe for future analysis.

## Limitations of the Data and Model

One limitation of this study is the incomplete dataset, with some states missing from the polling data. This limits our ability to produce a fully accurate prediction, as certain areas are excluded from the model. Additionally, the focus on state-level data may overlook finer voting patterns within states, particularly in districts with unique political identities.

## Future Research Directions

Future studies could improve by incorporating demographic variables such as age, income, and party affiliation, which might offer a clearer understanding of voter preferences. Examining additional state-level policies, like healthcare or education, may also provide insights into factors shaping voter choices.

Another possible enhancement involves increasing the geographic specificity of the model to include congressional districts or smaller regions, which may lead to more precise forecasts, especially in competitive states. As polling data collection methods evolve, combining traditional regression with machine learning techniques could allow for models that adapt to ongoing changes in public opinion, resulting in more timely and adaptable forecasts.

>>>>

### Percent support for Donald Trump
The primary outcome variable is the percentage (`pct`) of respondents who indicated support for Donald Trump in the 2024 general election. This percentage is calculated based on the total number of respondents in each poll divide by `sample_size`. The data includes observations from various states, providing both national and state-specific measures of Trump’s support.

@@

### Polling methodology
The `methodology` variable describes the method each pollster used to collect data. It includes approaches such as live phone interviews, online surveys, and mixed methods. For polls that used multiple collection methods (e.g., phone and online), we labeled them as "Mixed Voting" for simplicity. The methodology is important because different approaches can introduce varying degrees of bias, influencing the final reported percentages.

@@

# Define the number of electoral votes for each state
electoral_seats <- data.frame(
  state = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida",
            "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", 
            "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", 
            "Missouri", "Montana", "Nebraska", "Nevada", 
            "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
            "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", 
            "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming", 
            "District of Columbia"),
  electoral_votes = c(9, 3, 11, 6, 54, 10, 7, 3, 30, 16, 4, 4, 19, 11, 6, 6, 8, 8, 4, 10, 11, 15, 10, 6, 10, 
                      4, 5, 6, 4, 14, 5, 28, 16, 3, 17, 7, 8, 19, 4, 9, 3, 11, 40, 6, 3, 13, 12, 4, 10, 3, 3)
)

@@

### Polling score and transparency
The `numeric_grade` variable is a rating assigned to each polling organization, indicating the reliability of the poll. Pollsters with higher numeric grades are considered more reliable and consistent in their methodology (e.g. higher than 2.5). Additionally, the `transparency_score` measures how openly a pollster shares their methodology and data, which further affects the trustworthiness of their results.

@@

- **Polling scores**: Most organizations have numeric grades between 2.8 and 3.0, indicating that the majority of pollsters are considered highly reliable. Very few have grades below 2.7, which would suggest lower reliability.

- **Transparency scores**: Transparency scores show more variability, with many pollsters achieving scores above 9, reflecting a high level of openness in their methodology. Fewer organizations have transparency scores below 6, indicating that while some pollsters are less open, the majority strive for transparency in their practices.

### Sample size
The `sample_size` variable represents the total number of respondents surveyed in each poll. A larger sample size generally leads to more reliable and precise estimates of voter sentiment, as it reduces the margin of error. In contrast, smaller sample sizes can result in greater variability and less confidence in the results.

### Poll duration
The `duration` variable is a constructed variable that represents how long Donald Trump has been in the 2024 presidential campaign, measured as the number of days from his official campaign announcement on July 15, 2024 until the `end_date` of each poll. This variable helps quantify how much time has passed since Trump officially entered the race and allows us to examine how his support has evolved over the course of his campaign.

# Model {#sec-model}

Our modeling approach aims to quantify the relationship between various polling metrics and the percentage support for Donald Trump. For this analysis, we use a linear model to examine how factors such as poll score, methodology, poll duration, sample size, transparency score, and state influence support percentages. The model is implemented using the `stan_lm` function, with a Gaussian distribution to capture the variability in support rates.

@@

The model assumes that the distribution of support percentage, given these polling characteristics, follows a normal distribution. This Gaussian assumption facilitates parameter estimation, which is a standard approach in linear regression. To prevent overfitting and ensure interpretability, we assume moderate priors, maintaining balanced uncertainty across our predictors. This approach enables us to assess the impact of polling characteristics on candidate support while ensuring stability in our findings. Background details and diagnostics are included in the appendix.

## Alternative model

Initially, we attempted to categorize the `state` variable into two levels: state-specific polls and national polls. However, after modeling, we found that this approach was less effective for predicting election outcomes on a state level. Since the election is determined by the state-by-state electoral college, it became clear that using a state-specific variable would more accurately capture the variation needed to predict outcomes by state. This refinement allowed the model to better align with the electoral structure, resulting in a more accurate reflection of support distribution across states.

## Model set-up

The model predicts Trump's support percentage using the following predictor variables:

- Poll score (`numeric_grade`): Represents the quality rating of the poll.

@@

- Poll duration (`duration`): The the length of the election period was conducted.

- Sample size (`sample_size`): The number of respondents in the poll.

- Transparency score (`transparency_score`): A measure of how transparent the polling data and methodology are.

@@

\begin{align*}
    y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
    \mu_i &= \beta_0 + \beta_1 \cdot \text{Poll score}_i + \beta_2 \cdot \text{Methodology}_i \\
          &\quad + \beta_3 \cdot \text{Poll duration}_i + \beta_4 \cdot \text{Sample size}_i \\
          &\quad + \beta_5 \cdot \text{Transparency score}_i + \beta_6 \cdot \text{State}_i + \epsilon_i \\
    \epsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align*}

@@

### Model justification

Existing research and political science theories suggest that factors such as poll quality, sample size, transparency, methodology, poll duration, and the state where the poll takes place can influence Trump's support percentage. Higher quality polls and larger sample sizes are believed to yield more reliable estimates, while transparency scores can affect levels of public trust, potentially altering support. Moreover, different polling methodologies (e.g., online surveys, telephone interviews) may affect how representative the sample is, while longer polling duration can reflect shifts in public sentiment over time. State-specific factors, such as local political events or population characteristics, can also contribute to the variation in support. The timing of the poll, particularly in relation to key political events, might further influence public opinion.

@@

We predicted the election outcomes based on our test dataset. However, due to missing data for certain states in our original dataset, predictions could not be generated for all states. To address this gap, we reviewed historical trends from recent elections involving Joe Biden; if a state consistently leaned toward the same candidate, we assumed this trend would continue and used it to estimate outcomes for those missing states. This limitation arises because the states with NA values in the dataset are not represented in the model, preventing accurate forecasts for those regions. Further discussion on this issue is provided in @sec-discussion.

## Model results and interpretation

The linear regression model built by our training dataset, using 242 data points, estimated the factors influencing Trump's support levels. For brevity, @tbl-summary-model only shows the first ten rows of the model’s coefficients, with the full model summary available in the appendix. The intercept is estimated at 44.224, representing the baseline support level when all predictors are at their reference levels. The model achieved an $R^2$ value of 0.803, indicating that 80.3% of the variance in polling outcomes is explained by the predictors included in the model. The adjusted $R^2$ value of 0.753 confirms that the model captures the relationships between variables without significant overfitting.

```{r}
#| label: tbl-summary-model
#| tbl-cap: "Summary of key coefficients from the linear regression model predicting polling outcomes"
#| echo: false
#| warning: false

# Load the model
model <- readRDS(file = here::here("models/first_model.rds"))

summary <- summary(model)

summary_table <- head(summary, 10)[, 1:1]

kable(col.names = c(" ", "coefficient"),
  summary_table, digits = 3)
```

Certain predictors have a more obvious impact than others. For example, `numeric_grade` (0.779) shows that a one-unit increase in grade corresponds to a 0.779 percentage point increase in predicted polling outcomes. The type of methodology also plays a important role: "Mixed Voting" (1.692) and "Online Ad" (0.960) methods are associated with positive effects, while "Online Panel" (-1.037) and "Probability Panel" (-1.035) methods have negative effects relative to the baseline. Additionally, the state-specific coefficient shows a substantial effect. This underlines the importance of geographic context in polling predictions, especially for politically distinct areas.

Other variables, like `transparency_score` (-0.220) and `duration` (0.028), have smaller effects. The negative coefficient for `transparency_score` suggests a slight decrease in predicted support as transparency increases, potentially reflecting a sensitivity among the public to the transparency level in polling methods. Meanwhile, `sample_size` (0.002) shows a small positive effect, indicating that larger sample sizes modestly increase predicted support. A complete overview of all predictors is provided in the appendix.

## Predicted electoral outcomes

We employed a regression model to estimate the percentage of votes Trump is likely to receive in each state. Using these predicted results and the electoral vote allocations, we identified the winner for each state and calculated the total electoral votes for both Trump and Harris.

@@

- In Arizona, Trump is expected to receive 48.94% of the vote, resulting in a win for Harris with 11 electoral votes.
- In Texas, Trump is projected to receive 51.83%, thereby winning all 40 electoral votes for that state.

@@

# Define the number of electoral votes for each state
electoral_votes <- data.frame(
  state = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware", "Florida",
            "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky", "Louisiana", "Maine", 
            "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", 
            "Missouri", "Montana", "Nebraska", "Nevada", 
            "New Hampshire", "New Jersey", "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
            "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina", "South Dakota", "Tennessee", 
            "Texas", "Utah", "Vermont", "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming", 
            "District of Columbia"),
  electoral_votes = c(9, 3, 11, 6, 54, 10, 7, 3, 30, 16, 4, 4, 19, 11, 6, 6, 8, 8, 4, 10, 11, 15, 10, 6, 10, 
                      4, 5, 6, 4, 14, 5, 28, 16, 3, 17, 7, 8, 19, 4, 9, 3, 11, 40, 6, 3, 13, 12, 4, 10, 3, 3)
)

# Use the trained regression model to predict Trump's support percentage in each state
analysis_data_test$predicted_percent_trump <- predict(model, newdata = analysis_data_test)
total_trump <- aggregate(predicted_percent_trump ~ state, data = analysis_data_test, FUN = mean)

@@

# Create and display the results table
kable(state_predictions[, c("state", "predicted_percent_trump", "electoral_votes", "winner")], 
      col.names = c("State", "Trump Predicted %", "Electoral Votes", "Winner"), 
      digits = 2)
```

Since we are missing data for some states in our dataset, we turned to historical trends from recent elections involving Joe Biden to fill these gaps. For each missing state, we assessed whether it consistently leaned toward a particular candidate in past elections. If so, we assumed this trend would persist and used it as a basis for estimating outcomes in the current election.

@tbl-history summarizes the outcomes for the missing states based on this approach. It shows the predicted winner for each state based on historical trend. For instance, Alabama, Alaska, Arkansas, Idaho, Indiana, and Iowa are expected to go to Trump, contributing their respective electoral votes to his total. Conversely, states like Colorado, Connecticut, Delaware, Hawaii, and Illinois are projected to favor Harris, thus allocating their electoral votes to her.

```{r}
#| label: tbl-history
#| tbl-cap: Estimated Electoral Outcomes for Select U.S. States Based on Historical Voting Trends
#| echo: false
#| warning: false
#| message: false

all_states = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware",
                        "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
                        "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", 
                        "Missouri", "Montana", "Nebraska", "Nevada","New Hampshire", "New Jersey", "New Mexico", "New York",
                        "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", 
                        "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", 
                        "West Virginia", "Wisconsin", "Wyoming", "District of Columbia")

uncommon_states <- setdiff(all_states, common_states)

# Map the electoral votes to the uncommon states
uncommon_states_votes <- electoral_votes %>%
  filter(state %in% uncommon_states)

missing_states <- data.frame(uncommon_states_votes,
                             winner = c("Trump", "Trump", "Trump", "Harris", "Harris",
                                          "Harris", "Harris", "Trump", "Harris", "Trump", "Trump",
                                          "Trump", "Trump", "Trump", "Trump", "Trump",
                                          "Trump", "Harris", "Trump", "Trump", "Harris", "Trump", "Trump", 
                                          "Trump", "Trump", "Harris", "Harris", "Trump", "Trump", "Harris"))

# Create and display the results table
kable(missing_states, 
      col.names = c("State", "Electoral Votes", "Winner"), 
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  scroll_box(width = "100%", height = "400px")
```

Now, we combine the results from our predicted data and the historically inferred estimates for missing states. This merged dataset provides a view of the expected electoral outcomes across all states. @tbl-final-prediction presents the final predictions, showing the estimated electoral votes and projected winner for each state based on both model predictions and historical trends.

```{r}
#| label: tbl-final-prediction
#| tbl-cap: Final Predicted Electoral Outcomes by State, Combining Model Estimates and Historical Trends for Missing Data
#| echo: false
#| warning: false
#| message: false

# Remove the predicted_percent_trump column from state_predictions to match missing_states
state_predictions <- state_predictions %>%
  select(state, electoral_votes, winner)

# Merge the two datasets
final_predictions <- full_join(state_predictions, missing_states, by = c("state", "electoral_votes", "winner"))

# Arrange the rows by state name
final_predictions <- final_predictions %>%
  arrange(state)

# Display the final table
kable(final_predictions, 
      col.names = c("State", "Electoral Votes", "Winner"), 
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  scroll_box(width = "100%", height = "400px")
```

## Final electoral outcomes

Based on our analysis, @tbl-final-outcome summarizes the total predicted electoral votes for Trump and Harris in the 2024 U.S. Presidential Election. In the United States, a candidate must secure at least 270 electoral votes to win the presidency. According to these predictions, Kamala Harris is expected to secure the majority with 341 electoral votes, while Donald Trump is projected to receive 197. This suggests a clear victory for Harris.

```{r}
#| label: tbl-final-outcome
#| tbl-cap: "Predicted Final Electoral Votes for the 2024 U.S. Presidential Election: Trump vs. Harris"
#| echo: false
#| warning: false
#| message: false

# Calculate the electoral votes for each candidate
final_predictions$trump_electoral_votes <- ifelse(final_predictions$winner == "Trump", final_predictions$electoral_votes, 0)
final_predictions$harris_electoral_votes <- ifelse(final_predictions$winner == "Harris", final_predictions$electoral_votes, 0)

# Summarize the total electoral votes for each candidate
trump_electoral_votes <- sum(final_predictions$trump_electoral_votes)
harris_electoral_votes <- sum(final_predictions$harris_electoral_votes)

final_outcome <- data.frame(
  Candidate = c("Trump", "Harris"),
  `Final Votes` = c(trump_electoral_votes, harris_electoral_votes)
)

# Display the final table with added styling
kable(final_outcome, col.names = c("Candidate", "Final Votes"), digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  row_spec(1, background = "#FFCCCC") %>%    # Add light orange background for Trump row
  row_spec(2, background = "#C1D7FF") %>%    # Add light blue background for Harris row
  column_spec(2, bold = TRUE)
```

```{r}
#| label: fig-statemap
#| fig-cap: "The predicted winner for each U.S. state in the 2024 presidential election based on our model. States shaded in blue are projected to favor Harris, while those in red are projected to favor Trump."
#| echo: false
#| warning: false
#| message: false
# Add color information based on the winner
final_predictions <- final_predictions %>%
  mutate(winner_color = ifelse(winner == "Trump", "#FF6F61", "#6495ED"))

# Plot the map using usmap and ggplot2
plot_usmap(data = final_predictions, values = "winner_color", regions = "states") +
  scale_fill_identity(
    name = "Predicted Winner",
    labels = c("Harris", "Trump"),
    guide = "legend"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 18),
    plot.background = element_rect(fill = "white", color = NA)
  )
```

@fig-statemap displays the predicted outcome for each state in the 2024 U.S. Presidential Election, with red representing states where Trump is expected to win and blue for those favoring Harris.

It shows Trump gaining support across central and southern states, including Texas and Missouri, which have traditionally leaned Republican. Harris, meanwhile, is projected to lead in states along the West Coast and in the Northeast, such as California and New York, regions with a history of Democratic support. This visual representation complements the previous electoral vote projection, where Harris is anticipated to surpass the 270-vote threshold needed to win the presidency.

@@

## Interpretation of polling data and electoral implications

This paper examines the factors affecting Donald Trump’s polling performance in the 2024 U.S. Presidential Election. Our linear regression model shows that numeric grade, polling methodology, and state-specific factors are key predictors of Trump’s support. Higher numeric grades, representing perceived poll reliability, are positively associated with Trump’s polling numbers, suggesting that well-regarded polls may capture a more favorable sentiment. Polling methodology also plays a significant role, with different methods (e.g., "Mixed Voting" or "Online Panel") showing varied effects on the results, likely reflecting differences in respondent engagement or accessibility. State-specific factors underscore the importance of regional dynamics. The Electoral College system makes state-level support crucial, and unique local issues and demographics in each state contribute to Trump’s polling outcomes.

Transparency score and sample size also affect polling reliability. Polls with higher transparency and larger sample sizes provide more stable projections. While these factors improve data quality, they have less direct influence on voter preferences than state and methodology variables.

In contrast, poll duration has minimal impact in our model, suggesting that short-term and long-term polls yield similar support levels in this context. This finding implies that voter sentiment is more influenced by geographic and methodological context than by the length of polling periods. Overall, this model underscores the importance of state-level factors and methodology over specific polling practices in understanding Trump’s support trajectory in this election.

## Influence of state policies on voter preferences

Beyond the mechanics of polling, our analysis suggests that state-specific policies and regional cultural factors could influence voter preferences and eventually shape electoral outcomes, although they are not directly measured in our model. For example, state policies on recreational marijuana legalization may reflect broader ideological values that align with particular candidates or parties. To explore this idea, we compared our predictions map with a visualization of states where recreational marijuana is legal.

@fig-statemap-factor shows states where recreational marijuana is legal. Compared with @fig-statemap, many states that have legalized marijuana, such as California, Washington, and New York, also appear in our prediction as likely to support Harris, suggesting a correlation between progressive state policies and support for more liberal candidates. This policy environment may align with Democratic platforms on social issues, which could explain stronger support for Harris in these regions.

In contrast, states that prohibit recreational marijuana use—many of which are predicted to lean toward Trump—might favor more conservative social policies, possibly influencing their political alignment. This could reflect a cultural environment that aligns more closely with Republican stances, contributing to stronger support for Trump in these areas.

While the connection between marijuana legalization and voter preferences is not absolute, the alignment of state policies with broader political values suggests that cultural factors may influence voter preferences. Additionally, policies related to economic investment, healthcare, or education could shape these patterns, offering further influences into state-level political dynamics.

```{r}
#| label: fig-statemap-factor
#| fig-cap: "The legal status of recreational marijuana use in each U.S. state as of 2024. States shaded in green have legalized recreational marijuana, while those in gray have not."
#| echo: false
#| warning: false
#| message: false

legal_marijuana_states <- c(
  "Colorado", "Washington", "Alaska", "Oregon", "District of Columbia", "California",
  "Maine", "Massachusetts", "Nevada", "Michigan", "Vermont", "Guam", "Illinois", 
  "Arizona", "Montana", "New Jersey", "New York", "Virginia", "New Mexico", 
  "Connecticut", "Rhode Island", "Maryland", "Missouri", "Delaware", "Minnesota", "Ohio"
)

all_states = c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware",
                        "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
                        "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", 
                        "Missouri", "Montana", "Nebraska", "Nevada","New Hampshire", "New Jersey", "New Mexico", "New York",
                        "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", 
                        "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", 
                        "West Virginia", "Wisconsin", "Wyoming", "District of Columbia")
